Inference and Synthesis of\\ co-dimension one Bifurcations

### 2020-07-20 docs/article.tex
adding section on region velocity

Figure \ref{fig:tangent-field} shows how the bifurcation curve defined by $\rates(z)=0$ picks out one of many traces in tangent field $\tangent(z)$ for the saddle and pitchfork. The tangent field $\tangent(z)$ can always be analytically evaluated by taking the determinant in \eqref{eq:tangent-field}. It becomes practical to proceed with calculations on $\tangent(z)$ in the whole space $z$ and pick out a single trace by solving $\rates(z)=0$ later. For the saddle-node and pitchfork respectively
\begin{align}
    \tangent(z)=(\,3\theta_2 u^2-\theta_1\,)\,\hat{p}+\hat{u}
    \qquad\qquad
    \tangent(z)=(\,3\theta_2 u^2-p\,)\,\hat{p}+u\hat{u}
    \label{eq:tangent-field-examples}
\end{align}

@@

Consequently the curvature of the determinant $\partial_{\tangent}^2\Det$ along the tangent field $\tangent(z)$ in the vicinity of bifurcations tends to increase. Fortunately this curvature can always be obtained in closed analytical form. The second directional derivative with respect to tangent field $\tangent(z)$
\begin{align}
    \partial_{\tangent}^2\Det &:=
    \bigg(
        \partial_z\left(
            \partial_z\Det \cdot \hat{\tangent}
        \right)
    \bigg)\cdot \hat{\tangent}\quad
    \mathrm{where} \quad \hat{\tangent}:=\frac{\tangent}{|\tangent|}\\
    &=
    \hat{\tangent}\cdot\partial_z^2\Det\cdot\hat{\tangent}
    \quad+\quad
    \partial_z\Det\cdot\partial_z\hat{\tangent}\cdot\hat{\tangent}
\end{align}
The first term in the expression above is the usual result involving the Hessian matrix $\partial_z^2\Det$ when taking second order directional derivatives. The second term involving the gradient $\partial_z\Det$ appears because $\hat{\tangent}(z)$ is a function of $z$ giving rise to a non-zero Jacobian matrix $\partial_z \hat{\tangent}$. Using expressions for tangent fields \eqref{eq:tangent-field-examples} and determinants for the saddle-node and pitchfork models the determinant curvatures respectively become
\begin{align}
    \partial_{\tangent}^2\Det =
    -\frac{6 \theta_2 \left(\theta _1^2-9 \theta _2^2 u^4+2\right)}
    {\left(\left(\theta _1+3 \theta _2 u^2\right){}^2+2\right){}^2}
    \qquad
    \partial_{\tangent}^2\Det =
    -\frac{2 u^2 \left(p+3 \theta _2 u^2\right) \left(9 \theta _2 \left(p+\theta _2 u^2\right)+2\right)}{\left(p^2+6 \theta _2 p u^2+9 \theta _2^2 u^4+2 u^2\right){}^2}
\end{align}
\subsection{Semi-supervised Cost Function}

This curvature of the determinant $\partial_{\tangent}^2\Det$ can be extremised to encourage bifurcations. We can treat this curvature as an unsupervised correction of magnitude $\lambda$ to a supervised likelihood $L_{\theta}(z|\targets)$. For a particular bifurcation curve defined by $\rates(z)=0$ we can express the cost function as the negative log-likelihood
\begin{align}
    \loss(\theta|\targets) := -\log\left(\int_{\rates(z)=0}\!
    L_{\theta}(z|\targets) + \lambda\left( \frac{\partial^2\Det}{\partial\tangent^2} \right)^2
    \,\mathrm{d}z\right)\\
    L_{\theta}(z|\targets) := \mathbb{e}^{-|\targets-p|^2-\Det(z)^2}\qquad\qquad\quad
\end{align}
where we have chosen a Gaussian likelihood to bias optimisation of points in $z$ where the determinant $\Det(z)$ is close to zero towards targets $\targets$. Here we note that the integrand as well as the limits are functions of $\theta$ and $z$.

>>>>

Figure \ref{fig:tangent-field} shows how the bifurcation curve defined by $\rates(z)=0$ picks out one of many traces in tangent field $\tangent(z)$ for the saddle and pitchfork. The tangent field $\tangent(z)$ can always be analytically evaluated by taking the determinant in \eqref{eq:tangent-field}. It becomes practical to proceed with calculations on $\tangent(z)$ in the whole space $z$ and pick out a single trace by solving $\rates(z)=0$ later. For the saddle-node and pitchfork respectively
\begin{align}
    \tangent(z)=\hat{u}-(\,3\theta_2 u^2+\theta_1\,)\,\hat{p}
    \qquad\qquad
    \tangent(z)=u\hat{u}-(\,3\theta_2 u^2+p\,)\,\hat{p}
    \label{eq:tangent-field-examples}
\end{align}

@@

Consequently the curvature of the determinant $\partial_{\tangent}^2\Det$ along the tangent field $\tangent(z)$ in the vicinity of bifurcations tends to increase. Fortunately this curvature can always be obtained in closed analytical form. The second directional derivative with respect to tangent field $\tangent(z)$
\begin{align}
    \partial_{\tangent}^2\Det &:=
    \bigg(
        \partial_z\left(
            \partial_z\Det \cdot \hat{\tangent}
        \right)
    \bigg)\cdot \hat{\tangent}\quad
    \mathrm{where} \quad \hat{\tangent}:=\frac{\tangent}{|\tangent|}\\
    &=
    \hat{\tangent}\cdot\partial_z^2\Det\cdot\hat{\tangent}
    \quad+\quad
    \partial_z\Det\cdot\partial_z\hat{\tangent}\cdot\hat{\tangent}
\end{align}
The first term in the expression above is the usual result involving the Hessian matrix $\partial_z^2\Det$ when taking second order directional derivatives. The second term involving the gradient $\partial_z\Det$ appears because $\hat{\tangent}(z)$ is a function of $z$ giving rise to a non-zero Jacobian matrix $\partial_z \hat{\tangent}$. Using expressions for tangent fields \eqref{eq:tangent-field-examples} and determinants for the saddle-node and pitchfork models the determinant curvatures respectively become
\begin{align}
    \partial_{\tangent}^2\Det =
    -\frac{6 \theta_2 \left(\theta _1^2-9 \theta _2^2 u^4+1\right)}
    {\left(\left(\theta _1+3 \theta _2 u^2\right){}^2+1\right){}^2}
    \qquad
    \partial_{\tangent}^2\Det =
    -\frac{2 u^2 \left(p+3 \theta _2 u^2\right) \left(9 \theta _2 \left(p+\theta _2 u^2\right)+1\right)}{
    ( \left(p+3 \theta _2 u^2\right)^2+u^2 )^2
    }
\end{align}
\subsection{Semi-supervised Cost Function}

This curvature of the determinant $\partial_{\tangent}^2\Det$ can be extremised to encourage bifurcations. We can treat this curvature as an unsupervised correction of magnitude $\lambda$ to a supervised likelihood $L_{\theta}(z|\targets)$. For a particular bifurcation curve defined by $\rates(z)=0$ we can express the cost function as the negative log-likelihood
\begin{align}
    \loss(\theta|\targets) := -\log\left(\int_{\rates(z)=0}\!
    L_{\theta}(z|\targets) + \lambda\,K_{\theta}(z)^2\,
    \,\mathrm{d}z\right)\\
    L_{\theta}(z|\targets) := \mathbb{e}^{-|\targets-p|^2/\varepsilon-\Det(z)^2}\qquad
     K_{\theta}(z):=\frac{\partial^2\Det}{\partial\tangent^2}
\end{align}
where we have chosen a Gaussian likelihood to bias optimisation of points in $z$ where the determinant $\Det(z)$ is close to zero towards targets $\targets$. Figure \ref{fig:likelihood} shows how the likelihood $L_{\theta}(z|\targets)$ is sharply peaked around the bifurcation points $\predictions(\theta)$ and is maximal when they match the targets. The parameter $\varepsilon$ controls the desired precision of the likelihood function and $\lambda$ controls the strength of the unsupervised bias.

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=7cm]{likelihood-1.pdf}
\includegraphics[width=7cm]{likelihood-2.pdf}
\caption{Likelihood $L_{\theta}(z|\targets)$ shown in dark gold is sharply peaked around the bifurcation points $\predictions(\theta)$ and is maximal when they match the targets (Right). The light gold envelope shows the maxima of the likelihood as the points $\predictions(\theta)$ sweep across different values of $p$}
\label{fig:likelihood}
\end{figure}
Applying the gradient operator with respect to $\theta$ to the cost function, we find that not only the integrand but also the integration region needs to be differentiated. This requires the general Leibniz rule \cite{Flanders1973DifferentiationSign} which yields
\begin{align}
\frac{\partial\loss(\theta|\targets)}{\partial\theta}=
-\frac{
\int_{F=0}
  \mathrm{div}\!
\left(\frac{\partial z}{\partial\theta}
  \left(L+\lambda K^2\right)
\right)
+\frac{\partial L}{\partial\theta}+
2\lambda K\frac{\partial K}{\partial\theta}
\,\,\mathrm{d}z
}{
\int_{F=0}
\left(
  L+\lambda K^2
\right)\,\mathrm{d}z
}\label{eq:gradient}
\end{align}
where $\frac{\partial z}{\partial\theta}$ is the velocity field of the integration region due to changes in $\theta$. Fortunately the velocity field of implicitly defined regions can be solved for analytically \cite{Jos2011OnSurface}. This is done by taking the total derivative of the implicit equation
\begin{align}
    \frac{d\rates(z)}{d\theta}=\frac{\partial F}{\partial\theta}+
    \frac{\partial F}{\partial z}\frac{\partial z}{\partial\theta}
\end{align}
We can rearrange for $\frac{\partial z}{\partial\theta}$ using the Moore-Penrose inverse of the augmented Jacobian matrix $\frac{\partial F}{\partial z}$ which appeared in equation \eqref{eq:tangent-field}. Since the surface is defined by $\rates(z)=0$ we arrive at
\begin{align}
    \frac{\partial z}{\partial\theta} = - \frac{\partial F}{\partial z}^\top
    \left(\,
        \frac{\partial F}{\partial z}\,\frac{\partial F}{\partial z}^\top
    \right)^{-1}
    \frac{\partial F}{\partial\theta}
\end{align}
which for our two minimal models yield
\begin{align}
    \frac{\partial z}{\partial\theta} =
        \frac{ \hat{p}+\hat{u}\,(\,\theta_1+3\theta_2 u^2\,) }{1+(\,\theta_1+3\theta_2 u^2\,)^2}
        \left( u\,\hat{\theta}_1 + u^3\,\hat{\theta}_2 \right)
    \qquad
    \frac{\partial z}{\partial\theta} =
        \frac{ p\,\hat{u} + u\,(\,\hat{p}+3\theta_2u\,\hat{u}\,) }{(\,p+3\theta_2 u^2\,)^2 + u^2 }
        \left( \hat{\theta}_1 + u^3\,\hat{\theta}_2 \right)
    \label{eq:velocity-field-examples}
\end{align}
Thus the integrands in the gradient \eqref{eq:gradient} are completely analytic, leaving only region $\rates(z)=0$ to be evaluated numerically. Since these analytic forms can be pre-calculated, calculating the gradients shares the same computational complexity as calculating the steady states. 
\subsection{Basis Functions and Unsupervided Bias}
